How is experience transformed into memory?

### 2018-08-29 paper/main.tex
JRM finished pass through intro

\begin{abstract}
  How our experiences unfold over time define unique \textit{thought
    trajectories} through the relevant representational spaces.
  Within this geometric framework, one can compare the shape of the
  trajectory formed by an experience to that defined by our later
  remembering of that experience.  We propose a framework for mapping
  naturalistic experiences onto geometric spaces that characterize how
  they unfold over time.  New insights emerge when we apply this
  approach to a naturalistic memory experiment which had participants
  view and recount a video.  We found that the shapes of the
  trajectories formed by participants' recountings were all highly
  similar to that of the original video, but participants differed in
  the level of detail they remembered.  We also identified a network
  of brain structures that are sensitive to the ``shapes'' of our
  ongoing experiences, and an overlapping network that is sensitive to
  how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Removing an event from the context in which it occurrs often subtantially changes its meaning.  Second, the specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  Third, one might remember the gist or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.

How might one go about formally characterizing the ``gist'' of an experience?  Real world experiences unfold over extended time intervals.  Any given moment of an experience might derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g.\ ][]{LernEtal11}.  This interplay between our subjective experiences at different timescales defines a \textit{context} for each new moment, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Further, our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated).  This allows us to form stable estimates of our current situation and thereby behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  
%%%JRM STOPPED HERE
Although our experiences often change gradually, they occasionaly change suddenly~\citep[e.g., walking through a doorway][]{RadvZack17}.

Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure of the video, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This technique casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. In this framework, successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that closely follows the path (and covers the entirety) of the stimulus trajectory.

@@

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_bw.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participants' verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  \textbf{A.} A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. \textbf{B.} A sentences (294) by topics (100) matrix representing participant \#13's recall. \textbf{C.} A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The yellow boxes represent `events' recovered by a hidden Markov model. \textbf{D.} A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. \textbf{E.} An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  \textbf{F.} An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. \textbf{G.} A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. \textbf{H.} A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} \textbf{A.} 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. \textbf{B.} 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow. \textbf{C.} Individual trajectory plots for each participant overlayed with the video model (black). The direction of the recall sequence is indicated by the direction of arrows connecting each recall event.  The colors refer to the most likely video event.}
\label{fig:trajectory}
\end{figure}

@@

\subsubsection*{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} \textbf{A.} The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  \textbf{B.} The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} \textbf{A.} Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants \textbf{B.} Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g., the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. \textbf{C.} Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are individual recall events across all participants, where the colors correspond to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

% Supplemental materials
\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp1_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_matchmats.pdf}
\caption{\small \textbf{Video-recall event model correlation matrices.} Each participant's video event by recall event correlation matrix.  The yellow boxes represent the maximum correlation in each column.}
\label{fig:matchmats}
\end{suppfigure}

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp4_feature_value.pdf}
\caption{\small \textbf{Impact of individual features on topic modeling analysis.} \textbf{A.} Contribution of each feature to model structure. Bars represent the correlation of a video model trained in the absence of a given feature to the model trained on all features. \textbf{B.} Contribution of each feature to video model/recall model relationship. The leftmost bar represents the across-participants mean correlation between the video model and recall models trained on all features. Subsequent bars represent the same relationship between video and recall models trained in the absence of a given feature. Error bars are the standard error of the mean across participants.}
\label{fig:feature-importance}
\end{suppfigure}

@@

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp3_list_learning.pdf}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} A). The probability of first recall as a function of the serial position of the event during encoding. B). A lag-conditional response probability curve. Given recall of event i, the probability that the next recalled item will be from serial position i +/- lag. C). Proportion of events recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.}
\label{fig:list-learning}
\end{suppfigure}

>>>>

\begin{abstract}
How our experiences unfold over time define unique \textit{trajectories} through the relevant representational spaces.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  New insights emerge when we apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We found that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video, but participants differed in the level of detail they remembered.  We also identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Unlike in trial-based experiments, removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, precise recall is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the gist or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.  There is no analog of the ``gist'' of an experience in most traditional memory experiments; rather we tend to assess participants' abilities to recover specific details (e.g., the precise stimuli they encountered earlier in the experiment).

How might one go about formally characterizing the ``gist'' of an experience?  Any given moment of an experience might derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g.\ ][]{LernEtal11}.  Therefore perhaps the timecourse of how an event unfolds might be critical to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences typically change gradually, they also occasionally change suddenly~\citep[e.g., walking through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into compressed memories.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

Here we sought to examine how temporal dynamics of a ``naturalistic'' experience was reflected in participants' later memories of that experience.  We analyzed an open dataset which comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode (and of participants' verbal recalls).  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{topic trajectories} that describe how the experiences evolve over time.  In other words, the episode's topic trajectory is a formalization of its gist.  Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the gist of the original episode.  In addition, comparing the shapes of the topic trajectories of the original episode and of participants' retelling of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We also identified a network of brain structures whose responses (as participants watched the episode) reflected the gist of the episode, and a second network whose responses reflected how participants would later recount the episode.

%JRM STOPPED HERE

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_bw.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participants' verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  \textbf{A.} A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. \textbf{B.} A sentences (294) by topics (100) matrix representing participant \#13's recall. \textbf{C.} A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The yellow boxes represent `events' recovered by a hidden Markov model. \textbf{D.} A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. \textbf{E.} An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  \textbf{F.} An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. \textbf{G.} A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. \textbf{H.} A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} \textbf{A.} 2-dimensional embedding of the video events model (reduced with UMAP algorithm). The arrows indicate the forward direction of the video events. \textbf{B.} 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the group-average direction of all recall event transitions (i.e., a line segment connecting two consecutive recall event vectors) that intersected a circle (radius=.5) centered on the origin of the arrow. \textbf{C.} Individual trajectory plots for each participant overlayed with the video model (black). The direction of the recall sequence is indicated by the direction of arrows connecting each recall event.  The colors refer to the most likely video event.}
\label{fig:trajectory}
\end{figure}

@@

\subsubsection*{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} \textbf{A.} The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  \textbf{B.} The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} \textbf{A.} Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants \textbf{B.} Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g., the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. \textbf{C.} Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are individual recall events across all participants, where the colors correspond to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

% Supplemental materials
\begin{suppfigure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/supp1_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}

\begin{suppfigure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_matchmats.pdf}
\caption{\small \textbf{Video-recall event model correlation matrices.} Each participant's video event by recall event correlation matrix.  The yellow boxes represent the maximum correlation in each column.}
\label{fig:matchmats}
\end{suppfigure}

\begin{suppfigure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/supp4_feature_value.pdf}
\caption{\small \textbf{Impact of individual features on topic modeling analysis.} \textbf{A.} Contribution of each feature to model structure. Bars represent the correlation of a video model trained in the absence of a given feature to the model trained on all features. \textbf{B.} Contribution of each feature to video model/recall model relationship. The leftmost bar represents the across-participants mean correlation between the video model and recall models trained on all features. Subsequent bars represent the same relationship between video and recall models trained in the absence of a given feature. Error bars are the standard error of the mean across participants.}
\label{fig:feature-importance}
\end{suppfigure}

@@

\begin{suppfigure}[t]
\centering
\includegraphics[width=1\textwidth]{figs/supp3_list_learning.pdf}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} A). The probability of first recall as a function of the serial position of the event during encoding. B). A lag-conditional response probability curve. Given recall of event i, the probability that the next recalled item will be from serial position i +/- lag. C). Proportion of events recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.}
\label{fig:list-learning}
\end{suppfigure}
